---
title: null
output:
  html_document:
    code_download: true
    code_folding: show
    css: custom.css
    includes:
      in_header: header.html
      after_body: after_body.html
  pdf_document: default
---
<div class="content">

<h1 class="title">Modeling the Impact of Salary Distribution on NHL Team Success</h1>

<p class="subheading">
  A statistical examination using Gini coefficients and econometric models over 10 seasons in the National Hockey League
</p>

<h4>Abstract</h4>

<p class="abstract">
  This study investigates how intra-team salary distribution impacts team performance in the NHL, a league governed by a strict salary cap. Using the Gini coefficient, the Gini-squared, and Regulation + Overtime Wins (ROW) as the performance metric, we apply both a Poisson Generalized Linear Model (GLM) and a dynamic panel Generalized Method of Moments (GMM) estimator to determine the effect salary distribution has on a team's regular season wins. The results reveal a concave relationship between inequality and performance: teams with moderate salary dispersion tend to outperform those with either highly equal or highly unequal payrolls. Despite not accounting for outliers such as players on undervalued contracts or in-season trades, the findings offer a data-driven framework for optimizing roster construction under cap constraints.
</p>


```{r setup, include=FALSE}
library(readxl)
library(dplyr)
library(ggplot2)
library(ggfortify)
library(tibble)
library(tidyr)
library(pdynmc)
library(plm)

# Load Data
data <- read.csv("CompleteTeamData.csv")

# Add lagged ROW variable
data <- data %>%
  arrange(Team, Year) %>%
  group_by(Team) %>%
  mutate(ROW_prev_actual = lag(ROW)) %>%
  ungroup()
```

### Background

In salary-capped Professional Sports leagues like the National Hockey League (NHL), National Football League (NFL), and the National Basketball Association (NBA), where total team spending is constrained, performance disparities raise a fundamental question in analytics: how should teams allocate their fixed payroll to maximize success? While much of the focus has historically been on the effects of spending levels on outcomes, less attention has been paid to the internal distribution of salaries–how resources are spread between elite and depth forward rosters. The Gini coefficient, a standard measure of inequality, offers a unique statistical lens for quantifying these distributions. This study investigates whether there exists an optimal level of salary inequality that maximizes team performance.

Unlike some other sports leagues with more flexible spending rules, the NHL enforces a hard salary cap, meaning teams cannot exceed a fixed upper limit on total player compensation under any circumstances. Designed to ensure competitive balance across markets, the cap system drives franchises to make strategic trade-offs in building their rosters. Rather than simply outspending rivals, success in the NHL depends on how teams choose to structure their payroll within this rigid constraint–with the main challenge being finding balance between elite contracts, depth and durability across the lineup.

Each player’s contribution to the cap is measured by their cap hit, defined as the average annual value (AAV) of their contract, excluding performance bonuses. These cap hits are used to measure a team’s projected total cap hit, which can fluctuate due to factors like pro-rated contracts, buy-outs and long-term injured reserve (LTIR) placements. However, for the purpose of this study, these complex adjustments are excluded, and a team’s cap structure is simply defined as the sum of AAVs of players on its active roster.

Importantly, the cap ceiling (the maximum) and cap floor (the minimum) are set relative to a midpoint defined in the NHL’s Collective Bargaining Agreement. From 2019-2022, the league entered a “flat cap” period due to the pandemic, during which the cap ceiling remained static. This likely compressed salary distributions by reducing flexibility for high-end contracts and increasing the use of entry-level or league-minimum players–an important contextual factor for interpreting salary inequality over time and for calculating the Gini coefficient.

To evaluate whether salary inequality affects team outcomes, we needed a robust measure of performance that reflects team wide success in the salary-constrained environment of the regular season. This study uses Regulation + Overtime Wins (ROW) as the primary performance metric, obtained from the NHL’s official statistics database. ROW excludes shootout wins and better captures team-level skill and consistency over time.
These dynamics raise a natural question: is there an ideal balance in how teams distribute their payroll? Is it possible to leverage star power while maintaining roster depth to optimize regular season success under the cap?

This research investigates how the structure of player salary distribution within NHL teams influences overall team performance, independent of total team spending. Building on the framework established by Park (2022), who used the Gini coefficient to quantify intra-team salary inequality in professional sports, this study applies similar methods to NHL data. 

### Methods

To measure salary inequality within each team, this study uses the Gini coefficient, a widely accepted statistical measure of inequality dispersion that ranges from 0 (perfect equality) to 1 (perfect inequality). Originally developed for measuring economic income distributions, the Gini is applied here to the average annual values (AAV) of each NHL player’s contracts on a per-season, per-team basis. The coefficient has been calculated using the following formula:
\[
G = \frac{1}{n \sum_{i=1}^{n} x_i} \sum_{i=1}^{n} (2i - n - 1)x_i
\]
In this formula, \( n \) represents the number of players on a team’s roster, and \( x_i \) denotes the Average Annual Value (AAV) of the \( i \)-th player, where players are ranked from lowest to highest salary.

While the Gini coefficient provides a concise summary of intra-team inequality, it has several known limitations. It is sensitive to data quality, can return identical values for different distributions, and may be distorted in settings with bounded ranges—such as the NHL’s hard salary cap and floor. Nonetheless, it remains a valuable statistic for comparing how evenly teams allocate their payrolls across players each season.

In order to capture the non-linear effect of salary inequality on team performance, this study also includes the square of the Gini coefficient as an explanatory variable. The traditional use of Gini assumes a linear relationship between inequality and performance – suggesting that more or less inequality has a consistently positive or negative effect. However, in the NHL, where spending is bounded by cap ceilings and floors, this linear model is too restrictive. By incorporating $Gini^2$ in our model, the regression allows for curved relationships, which can capture the curvature since too much or too little inequality may both be bad. 
The updated performance model now becomes: 
\[
\text{Performance} = \beta_0 + \beta_1 \cdot Gini_i + \beta_2 \cdot Gini_i^2
\]
In this model, if \( \beta_1 > 0 \) and \( \beta_2 < 0 \), it suggests that performance increases with Gini up to a certain point before declining. Conversely, if \( \beta_1 < 0 \) and \( \beta_2 > 0 \), performance follows a U-shaped curve, indicating that both very low and very high inequality are disadvantageous for team performance.
The value of Gini that maximizes performance is estimated as:
\[
Gini_{\text{optimal}} = \frac{-\beta_1}{2\beta_2}
\]
Using both Gini and $Gini^2$ in the model is necessary since they capture different dimensions of the relationship: Gini measures the direction of the slope, while $Gini^2$ shows the curvature. Together, they allow the model to identify whether there exists an optimal level of inequality for maximizing team regulation and overtime wins while staying within the bounds of the salary cap. 

To model the relationship between salary distribution and team performance, this study applies a Poisson regression model with a log-link function–an appropriate Generalized Linear Model (GLM) for count data. Regulation + Overtime Wins (ROW), the dependent variable, is non-negative integer count, making the Poisson distribution a natural fit. We assume: 
\[
Y_i \sim \text{Poisson}(\lambda_i)
\]
where \( Y_i \) is the number of ROW wins for team \( i \) in a given season, and \( \lambda_i \) is the expected number of wins. The log-link function relates the predictors to the expected outcome via:

\[
\log(\lambda_i) = \beta_0 + \beta_1 Gini_i + \beta_2 Gini_i^2 + \beta_3 ROW_{i-1}
\]

This form ensures \( \lambda_i = \exp(\ast) > 0 \), satisfying the domain constraints of the Poisson distribution. The multiplicative nature of the covariates allows each coefficient to be interpreted in terms of percentage change for the expected number of wins. For instance, a one-unit increase in Gini would result in a multiplicative change of \( \exp(\beta_1) \) in \( \lambda_i \), all else held constant.

Three predictors are included in the model: the Gini coefficient to capture the linear effect of intra-team salary inequality, its square to account for potential nonlinear relationships, and previous season’s ROW to reflect autoregressive dynamics in team performance. This structure is motivated by the hypothesis that both very low and very high levels of inequality may be suboptimal, and that past performance can influence future outcomes. 

However, the inclusion of $ROW_{i-1}$ raises potential concerns about endogeneity. Team success in one season may be correlated with unobserved factors (e.g., coaching relationships, injuries, trades) that also affect the current season performance. 

To address potential endogeneity in lagged team performance and salary inequality measures, this study employs a linear dynamic panel Generalized Method of Moments (GMM) estimator using the `pdynmc` package. The model accounts for unobserved team-specific heterogeneity and high persistence in NHL Regulation + Overtime Wins. 
Linear dynamic panel data models are well-suited for capturing both temporal dynamics and unobserved heterogeneity. However, the inclusion of lagged dependent variables induces correlation with the error term, especially when fixed effects are present. As a result, standard estimation techniques such as ordinary least squares (OLS) or fixed-effects estimators yield biased and inconsistent estimates. GMM offers an alternative by using internally derived moment conditions to instrument for endogenous regressors. 
The following dynamic panel model is estimated: 
$$
ROW_{i,t} = \alpha ROW_{i,t-1} + \beta_1 Gini_{i,t} + \beta_2 Gini_{i,t}^2 + \eta_i + \varepsilon_{i,t}
$$

where team fixed effects $\eta_i$ are removed via first differencing. Lagged values of ROW (up to 7 lags) serve as instruments for the endogenous lagged dependent variable, and Gini terms are treated as endogenous with instruments drawn from their lagged values (up to lag 2).

Estimation is performed using two-step GMM with Windmeijer-corrected standard errors to address finite-sample bias. The instrument matrix is based on moment conditions in differences and excludes level and nonlinear conditions.

This setup follows the methodology of Arellano and Bond (1991), which is used in dynamic panel settings to address endogeneity from lagged dependent variables. Notably, this approach was also employed by Park (2022) in the original inspiration for this study. The implementation further reflects recommendations from Fritsch et al. (2021), specifically for handling persistence and improving finite-sample properties. 


While GMM offers a flexible way to address potential endogeneity, it produces more conservative estimates and can be sensitive to instrument weighting, which distorts predictor influence. In contrast, the Poisson GLM–based on maximum likelihood estimation–uses the correct variance and covariance structure for count data, yielding more efficient and stable estimates. Its ability to adjust for minor differences in ROW across teams make it a better overall fit for this study. 

### Results

##### Simulation Results: 
To assess the stability of coefficient estimates and explore how the Poisson GLM performs under varying salary inequality structures, we simulated team performance data using a log-link Poisson process. Coefficients are estimates from the observed data using a Poisson GLM, where Regulation + Overtime Wins (ROW) are modeled as a function of a second-order polynomial of the Gini coefficient and the lagged ROW. 
These fitted coefficients are then used to simulate new ROW values for each NHL team in each season. The simulation proceeds iteratively by team and year: the actual ROW is retained for each team’s first observed season, and for all subsequent years, expected \( ROW(\lambda_{it}) \) is calculated as:
\[
\log(\lambda_{it}) = \beta_0 + \beta_1 Gini_{i,t} + \beta_2 Gini_{i,t}^2 + \beta_3 ROW_{i,t-1}^{\text{sim}}
\]
ROW values are drawn from a Poisson distribution with mean \( \lambda_{it} \), introducing realistic randomness while preserving temporal dependence. After generating one complete dataset of simulated ROW values, the model is re-estimated using the same GLM specification. 
The simulation study, repeated 100 times using Gini coefficients from the real-data, produced stable and interpretable results that reinforce the hypothesized relationship between salary inequality and team performance. On average, the simulated intercept was approximately 3.11 with low variability (SD = 0.065), reflecting consistent baseline performance across teams. The mean coefficient was close to zero (0.004), suggesting mild curvature in the relationship. Despite variation across runs, the sign pattern is broadly consistent with the expected concave shape.
Boxplots and histograms of the coefficient distributions confirm that the intercept and Gini terms remained tightly clustered across replication, with no significant outlier, while the $Gini^2$ term exhibited greater variance, likely due to its nonlinear role and weaker signal. The autoregressive term for lagged ROW had a small positive mean (0.00175) with low standard deviation, indicating weak but stable persistence in performance over time. 
However, the individual simulation model (e.g., one run shown with Intercept = 3.15, $Gini$ = -1.17, $Gini^2$ = -0.04, LagROW = 0.0017) sometimes failed to reach significance on higher-order terms, illustrating noise introduced through Poisson sampling. 
Overall, the simulation results support the robustness checks of the GLM framework in recovering key relationships and demonstrate that moderate salary inequality tends to be connected to higher team performance–though the effect becomes imprecise at the extremes. 


GMM SIM WRITE UP HERE 

##### Real Results: 

The Poisson GLM was fitted to the actual team performance data using ROW as the outcome and a second-order polynomial of the Gini coefficient alongside lagged ROW as predictors. The model takes the form:

\[
\log(\lambda_i) = \beta_0 + \beta_1 Gini_i + \beta_2 Gini_i^2 + \beta_3 ROW_{i-1}
\]
All predictors are statistically significant at the 5% level. The linear Gini term is expected to be positive while the squared term is negative, indicating a concave relationship between salary distribution and team performance. This suggests that both too little and too much inequality can reduce a team’s ROW, with an optimal Gini range in between. The coefficient for lagged ROW was positive (0.0138, p<0.001), confirming that past performance is also a strong predictor of current success.
$$
\textit{Overdispersion} = \frac{\text{Residual Deviance}}{\text{Residual } df} = \frac{513.57}{266} \approx 1.93
$$
The residual deviance is higher than the degrees of freedom, indicating mild overdispersion. While not extreme, this suggests that slight variance inflation relative to the Poisson assumption. 

Model diagnostics indicate that the Poisson GLM provides a reasonably good fit for the observed data. The Residuals vs Fitted and Deviance Residuals vs Fitted plots reveal no clear patterns or systematic deviations, suggesting an appropriate mean-variance relationship under our model assumption. The Q-Q plot of standardized deviance residuals approximates normality, with only a few minor outliers at the distributional tails. The Scale-Location plot exhibits roughly homoscedastic variance across the range of fitted values, supporting the Poisson assumption of constant dispersion. 

```{r echo=TRUE, eval=TRUE}
#REAL GLM MODEL:
real_glm <- glm(
  ROW ~ poly(RawGini,2) + ROW_prev_actual,
  family = poisson(link = "log"),
  data = data %>% filter(!is.na(ROW_prev_actual))
)
summary(real_glm)
autoplot(real_glm, which = 1:6, ncol = 2) +
  theme_minimal(base_family = "serif") +
  theme(
    plot.background = element_rect(fill = "#F7FAFC", color = NA),
    panel.background = element_rect(fill = "#F7FAFC", color = NA),
    panel.grid.major = element_line(color = "#A8D8FF"),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "#A8D8FF", color = NA),
    strip.text = element_text(color = "#002244", size = 12, face = "bold"),
    axis.text = element_text(color = "#2D3748"),
    axis.title = element_text(color = "#002244", face = "bold"),
    plot.title = element_text(color = "#B22222", face = "bold", hjust = 0.5)
  )

cov2cor(vcov(real_glm))

anova(real_glm)
```

While the Residuals vs Leverage plot identifies a few data points with elevated leverage such as observations 239, 270 and 281–none show intense influence on the overall model fit, as confirmed by the Cook’s Distance plot. 

The estimated correlation matrix of the model coefficients shows low multicollinearity between predictors. The strongest correlation (-0.97) is between the intercept and the previous year’s ROW, which is expected due to the autoregressive structure. However, this is not problematic for interpretation given the significance and model stability. 

Collectively, these diagnostics support the reliability of the GLM estimates and suggest no substantial violations of the underlying assumptions. 
```{r echo=TRUE, eval=TRUE}
#DEVIANCE RESIDUALS: 
res <- residuals(real_glm, type = "deviance")
fitted_vals <- fitted(real_glm)
#Overdispersion: 
summary(real_glm)$deviance / summary(real_glm)$df.residual
#Influence:
cd <- cooks.distance(real_glm)
plot(cd,
     type = "h",
     lwd = 2,
     col = "#002244",       
     main = "Cook's Distance",
     xlab = "Observation",
     ylab = "Cook's Distance",
     col.main = "#B22222",  
     col.lab = "#002244",   
     col.axis = "#2D3748" 
)
top_n <- order(cd, decreasing = TRUE)[1:3]
text(x = top_n, y = cd[top_n], labels = top_n, pos = 3, col = "#B22222")

```
  
The results from the two-step GMM estimation provide strong support for a non-linear relationship between salary distribution and team performance in the NHL. The coefficients on the Gini coefficient and its square are both highly statistically significant and of opposite signs: ($\beta_1 = 1703$, $p < 0.001$; $\beta_2 = -2151$, $p < 0.001$) indicating a concave shape. This suggests that team success is maximized at a moderate level of salary inequality.. These findings closely mirror the results of Park (2022), who identified a similar hump-shaped relation in the NFL. 
Interestingly, the lagged ROW variable is not statistically significant in this model, implying that once fixed effects and endogeneity are accounted for, prior season performance has limited direct influence on current season performance. This may reflect roster changes and variability in NHL team dynamics. 
Model diagnostics reinforce the validity of the estimation. The Hansen’s J-test fails to reject the null hypothesis of instrument variability ($\chi^2 = 28.34$, $df = 46$, $p = 0.98$), and no signs of overfitting are detected.
While the F-statistic on the slope coefficient is not significant ($p = 0.53$), this is due to the insignificance of the lagged ROW term; the Gini variables remain individually and jointly robust.

### Interpretation
MODEL RESULTS HERE 

##### Optimal Gini Calculations: 
To translate the concave relationship between salary distribution and team performance into actionable insight, this study calculates the optimal Gini coefficient–the point at which team performance, as measured by Regulation + Overtime Wins, is maximized. Using the estimated coefficients from a Poisson regression with raw polynomial terms, the optimal Gini is: 
\[
Gini_{Optimal} = \frac{- \beta_1}{2 \beta_2}
\]
where ($\beta_1$) and ($\beta_2$) are the coefficients on the linear and squared Gini terms, respectively. This yields an optimal Gini value of approximately 0.408. While this estimate is not statistically significant in isolation, it offers a meaningful benchmark for NHL teams aiming to optimize regular season performance under salary cap constraints. The optimal Gini provides a practical tool for front offices to assess whether their current roster structure aligns with performance-maximizing patterns observed across the league. 
```{r echo=FALSE, eval=TRUE}
#OPTIMAL GINI CALCULATIONS: 
glm_raw <- glm(
  ROW ~ RawGini + Gini2 + ROW_prev_actual,
  family = poisson(link = "log"),
  data = data
)
summary(glm_raw)

beta1 <- coef(glm_raw)["RawGini"]
beta2 <- coef(glm_raw)["Gini2"]

Gini_optimal <- -beta1 / (2 * beta2)
Gini_optimal
```

These findings have important implications for how NHL teams should approach roster construction under the salary cap. The concave relationship between salary inequality and team performance suggests that neither a top-heavy roster loaded with superstars nor a fully equal distribution of salaries across all players maximizes success. Instead, the data indicates there is an optimal middle ground–teams that carefully balance high-salary contracts with cost-effective depth players tend to perform better in the regular season. 
For general managers, this means that pushing too far in either direction of distribution may undermine performance. The optimal Gini coefficient estimated in this study is within a range that is realistically attainable under current cap structures, meaning teams can use this insight to inform real-time roster planning. By monitoring their Gini trajectory across seasons and adjusting as needed, teams may gain a strategic edge without exceeding the constraints of the cap. 
It is important to note, however, that the model does not explicitly account for individual cases where players significantly outperform their contracts–such as young stars on entry-level contracts. These exceptions may allow teams to temporarily exceed expected performance at a given Gini level, but over time, roster-wide balance appears more predictive of success. 
Ultimately, the results provide a data-driven framework for evaluating trade-offs between star power and team cohesion, helping teams better navigate the rigid boundaries of the cap. 

### Conclusion

### Sources 

</div>
